---
title: "TTC Bus Delays"
author: "Timothius Prajogi"
date: today
format: html
bibliography: bibliography.bib
---

## Packages
- Base R [@citeR]
- Core `tidyverse` [@citetidyverse]
  - `dplyr` [@citedplyr]
  - `ggplot2` [@citeggplot]
- `janitor` [@citejanitor]
- `knitr` [@citeknitr]
- `opendatatoronto` [@citeopendatatoronto]

```{r}
#| message: false
#| warning: false

library("janitor")
library("knitr")
library("opendatatoronto")
library("tidyverse")
```

## The Data

```{r}
#| echo: false
#| message: false
#| warning: false



ttc_delay_data <- 
  read_csv(
    file = "../data/cleaned_ttc_bus_delay_statistics.csv",
    show_col_types = FALSE
  )

```


The cleaned dataset that we are working with:

```{r}
ttc_delay_data
```

## Graphs

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-delays
#| fig-cap: The staistics for the TTC in 2023.

ttc_delay_data |>
  ggplot(aes(x = incident)) + 
  geom_bar(stat = "count") +
  ggtitle("Amount of delays caused by incidents") + 
  xlab("Incident type") +
  ylab("Delay count")


```

We can see from @fig-delays that most of the delays were caused by bus issues. 
This includes of categories such as issues with the bus operator and mechanical
issues with the bus itself. 

The second most common cause of delays are safety issues, which covers sanitary
concerns, police and fire emergencies, and security issues like track
trespassing.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-time
#| fig-cap: The staistics for the TTC in 2023.


ttc_delay_data |>
  ggplot(aes(x = time)) +
  geom_line(stat = "count") + 
  ggtitle("Amount of delays as a function of time") + 
  xlab("Time of delay") +
  ylab("Delay count") +
  scale_x_time(breaks = scales::pretty_breaks(n = 12))


```

We can see from @fig-time that during non-peak operating hours are also the times
for non-peak delay times. As less buses run, understandably less delays occur.

Most of the delays occur during rush hour from around 4p.m. to 6p.m.. From 
@fig-delay we know that since most delays are caused by issues with the bus 
and/or the operator itself, meaning it is not because scheduling more 
buses at any time is more difficult, but just that there is a larger amount of
buses something could go wrong with. 
